Wildcard: End user modification of web applications through a data table

### 2019-12-29 paper.tex
intro

@@

\title{Wildcard: Modifying web applications by directly manipulating a
data table}
\subtitle{}% optional

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Browser extensions and user scripts can modify web applications in
useful ways, but many people have unique needs that aren't met by
existing extensions. Today, most of them are stuck---they can't build
their own browser extensions without learning how to program in
Javascript.

Wildcard is a platform that empowers anyone to build browser extensions
without doing traditional programming. Wildcard shows the main data from
a web page in a table with a bidirectional connection to the original
page. People can directly manipulate the table to sort/filter content,
add private annotations, and use spreadsheet formulas to fetch data from
other websites.
\end{abstract}

@@

Todos for the introduction:

\begin{itemize}
\tightlist
\item
  hook with Airbnb sort
\item
  add more bigger picture reasons, less focus on the affordances
\item
  make this shorter
\end{itemize}

People have complaints about web apps they use, but they rarely modify
those apps to meet their needs. Why not? Some guesses:

\begin{itemize}
\tightlist
\item
  Technical skill: Most people don't know how to use Javascript and
  manipulate the DOM.
\item
  Low ROI: It takes a long time to reverse engineer a site, and it's
  usually not worth it
\end{itemize}

These seem reasonable but they might not tell the whole story.

There have been many research projects that try to address these
hurdles. For example, Chickenfoot \autocite{bolin2005} allows for people
to more quickly modify sites without using Javascript or dealing with
the DOM, directly addressing both of these barriers. But, a decade
later, not many people have ended up using systems like this.

I know how to do web programming, and yet I rarely modify my apps.
Sometimes it actually turns out to be pretty easy to hack on a site once
I start doing it so the ROI is actually pretty high. These reasons don't
seem to fully explain my behavior.

A disheartening explanation might be that most people just don't care
enough to make changes. This claim might be true in the current context,
but it's also important to remember that motivation is connected to
culture and to the space of possibilities provided by our tools. Most
people probably didn't want to write letters before mass literacy.

One interesting explanation is that \textbf{people can't estimate the
difficulty of a change}. It's not motivating to think about what changes
I might want when I don't know whether they would take a few minutes or
are completely impossible. For me, an engineer, this gets at the heart
of the issue. I know I could probably implement the change, but it's so
hard to estimate how long it would take (minutes? months?) that I don't
even bother trying. In general, when something seems expensive to do (or
even possibly expensive), it can discourage casual lightweight
experimentation.

So, perhaps to encourage people to casually modify software,
\textbf{apps need to provide more consistent affordances indicating what
changes are possible and easy.} Maybe if it were more obvious that
certain types of changes could be achieved in mere minutes, programmers
and non-programmers would end up modifying our software more.

\hypertarget{the-wildcard-platform}{%
\subsubsection{The Wildcard platform}\label{the-wildcard-platform}}

Wildcard adds a panel to the bottom of a web page that shows a
structured table view of some of the core data in the page. When the
user manipulates the table, the original page also gets modified. We aim
to make the mapping between the table and the page as direct and
intuitive as possible.

Todos here:

\begin{itemize}
\tightlist
\item
  a better demo is actually sorting too at this point too, to show the
  bidirectional nature
\item
  hint at future possibilities
\item
  introduce an image fallback for the PDF version
\end{itemize}

@@

Wildcard is fairly general and can support many useful changes to
websites, which will be demoed later:

\begin{itemize}
\tightlist
\item
  sorting and filtering data: eg sorting shopping results
\item
  using 3rd-party APIs and performing small computations to add new
  data, in the style of ``web mashups'': eg adding walkability scores to
  hotel listings
\item
  adding private user annotations to the page: eg taking notes on
  different options
\item
  using alternate UI widgets to enter data into a page: eg using a
  personal datepicker widget with private calendar data, to enter the
  right dates for taking a flight
\end{itemize}

The overall goal is to provide generic tools that fit well with the
table paradigm and enable many specific useful changes. But it's
important to note that Wildcard doesn't aim to provide maximum coverage
all the possible ways someone might want to modify a web page. Rather,
it aims to provide a useful, simple subset of modifications, and to
provide consistent affordances so that users confidently understand
which modifications they can make.

@@

\hypertarget{implementation}{%
\section{Implementation}\label{implementation}}

Eventually web apps might provide the structured data table themselves.
In the meantime, we need some sort of adapter to make this system work
with existing sites.

Wildcard provides a system for creating a wrapper on top of existing
websites. This wrapper defines how structured data can be extracted out
of the page, and also how manipulating the table should modify the page.

The most basic way of building these wrappers is for skilled programmers
to manually build and maintain them for popular sites. This approach
beats the status quo because many people, including end users, can
benefit from the generic wrapper and use it in many ways. This is
different from the current world where programmers build
use-case-specific browser extensions, and each extension has to
implement its own interactions with the low-level DOM of a page. There's
also a greater incentive for many people to collectively maintain a
wrapper if it's shared.

A more advanced way would be to make these wrappers partially or totally
automated, and enable end users to create them. This future work could
leverage existing research on wrapper induction but isn't the focus of
the current work.

@@

Here are some of the ideas behind the design of Wildcard. We hope these
can be useful principles not only for Wildcard, but also for other tools
that enable end users to modify software.

@@

@@

In Wildcard, users don't interact with the original page to modify it;
instead they manipulate an alternate representation of the data in the
page. The tool uses visual cues to indicate the mapping between the
table representation and the original page. This is similar to the way
that browser developer tools use in-page highlighting to indicate the
mapping between HTML and the original page.

@@

We considered alternate approaches where the user would interact more
directly with the original page, but decided that the table view
provided a consistent

\begin{itemize}
\tightlist
\item
  True DM might be right on the page itself

  \begin{itemize}
  \tightlist
  \item
    There are good UIs for this, eg Chrome DevTools element picker
  \end{itemize}
\item
  Considered for Wildcard, but went with an indirect proxy. Still
  maintain a close mapping.
\item
  Pros: consistency across sites. Avoid site-specific styling issues.
\item
  Cons: maybe less intuitive? More work to do the mapping, might be a
  source of confusion.
\end{itemize}

\hypertarget{first-party-optional}{%
\subsection{First party optional}\label{first-party-optional}}

\begin{itemize}
\tightlist
\item
  3p-only is enough. 1p help is optional.
\item
  Can even do things the 1p didn't want to expose.
\item
  make it easy for 1p: just ask for a data table, no fancy schema.
\end{itemize}

\hypertarget{built-for-the-web}{%
\subsection{Built for the web}\label{built-for-the-web}}

\begin{itemize}
\tightlist
\item
  web has the right foundations:

  \begin{itemize}
  \tightlist
  \item
    the right platform
  \item
    open platform
  \item
    dev tools
  \item
    originally intended to be
  \end{itemize}
\item
  in-place toolchain: meet the user where they are, in the page
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  bring your own views of the data
\item
  bring your own widgets for data entry
\item
  instrumental interaction
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Instrumental interaction
\item
  Web automation
\item
  Wrapper induction
\item
  Personal data
\item
  Extension helpers
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  How far does functionality go?

  \begin{itemize}
  \tightlist
  \item
    workflows? triggers?
  \item
    what can and can't be done
  \end{itemize}
\item
  automated wrappers?

  \begin{itemize}
  \tightlist
  \item
    lean on existing tech
  \end{itemize}
\item
  usability studies
\end{itemize}

\hypertarget{meta-todos}{%
\section{meta: todos}\label{meta-todos}}

\begin{itemize}
\tightlist
\item
  remove bold text for the proceedings version
\item
  go through the Programming proceedings template checklist
\end{itemize}

>>>>

\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

\title{Wildcard: End user modification of web applications through a
data table}
\subtitle{}% optional

@@

% Please always include the abstract.
% The abstract MUST be written according to the directives stated in 
% http://programming-journal.org/submission/
% Failure to adhere to the abstract directives may result in the paper
% being returned to the authors.
\begin{abstract}
Browser extensions and user scripts can modify web applications in
useful ways, but many people have unique needs that aren't met by
existing extensions. Today, most of those people can't build their own
browser extensions, since they can't program in Javascript.

Wildcard is a platform that empowers anyone to build browser extensions
without doing traditional programming. Wildcard shows the main data from
a web page in a table that maintains a bidirectional connection to the
original page. By directly manipulating the table, people can
sort/filter content, add private annotations, and use spreadsheet
formulas to fetch data from other websites.
\end{abstract}

@@

In 2012, the travel site Airbnb removed the ability to sort listings by
price. Users could still filter down to a price range, but could no
longer view the cheapest listings first. Many users complained on online
message boards that the change seemed hostile to users---``It's so
frustrating!..What is the logic behind not having this function?'' says
one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

This is a familiar situation in a world of web applications that are
frequently updated without user consent. When web software doesn't quite
meet users' needs, their only recourse is to complain to the developers
and hope someone listens. Sometimes there is a way to use a browser
extension or user script to fix the issue, but making an extension
requires sophisticated programming skills. We see this as a waste of the
pliability of software and the openness of the Web. If users could make
tweaks to web applications without programming, a vast number of people
would gain the agency to modify software to meet their unique
preferences.

In this paper, we introduce a system called Wildcard\footnote{Wildcard
  was the original internal name for Hypercard, which pioneered both
  user modifiable software and the modern Web.} that aims to meet this
need. Wildcard adds a panel to the bottom of a web page that shows a
structured table view of some of the main data in the page. The table
maintains a bidirectional connection to the original page---when the
user manipulates the table, the original page also gets modified, and
vice versa.

In Wildcard, sorting Airbnb listings by price takes just one click on a
table header, with no programming required by the end user. Beyond
sorting and filtering data, Wildcard also supports a broad set of other
use cases like accessing third party APIs, performing small
computations, recording private user annotations to the page, and using
alternate UI widgets to enter data.

Wildcard currently requires a programmer to manually define an adapter
between the original page and the data table using web scraping
techniques. Even though programming is still required for part of the
process, this forms an effective collaboration between the programmer
and the end user, since the end user is able to perform a wide variety
of modifications to a site once a single adapter is defined. Eventually,
we envision adapter creation being less of a burden by making it
automated or making it possible for end users to build adapters.

Let's see an example of how an end user might modify a website with
Wildcard in practice.

@@

@@

\hypertarget{system-implementation}{%
\section{System Implementation}\label{system-implementation}}

\emph{Goal of this section: briefly explain the current implementation.
Enough detail to ground further discussion, but don't dwell on it. It's
not the main point, and will evolve a lot.}

\begin{itemize}
\tightlist
\item
  Built as a Greasemonkey script for now (\emph{todo: convert to a full
  broser extension?})
\item
  Describe the adapter API

  \begin{itemize}
  \tightlist
  \item
    programmer creates adapters for now
  \item
    maybe end users can create in the future, or automated
  \item
    show a snippet of adapter code
  \end{itemize}
\item
  Mention the technique of scraping data from AJAX requests
\item
  future implementation goals

  \begin{itemize}
  \tightlist
  \item
    make it easy for programmers to add adapters + plugins, and
    distribute them to users. (Currently all adapters + plugins are part
    of the main Wildcard codebase)
  \end{itemize}
\end{itemize}

@@

Below are some of the ideas behind the design of Wildcard. We hope that
these principles are also useful for thinking more broadly about
enabling end users to modify software.

@@

\emph{Todo: add a diagram here?}

@@

In Wildcard, users manipulate a data table added to the page, rather
than the original page. Although the interaction with the table is
direct like using a spreadsheet, the interaction with the page is
indirectly mediated through the table.

We considered other approaches where the user would interact more
directly with the original UI, e.g.~injecting sort controls into the
page, but decided that the consistency and affordances of the table view
outweighed the potential confusion of a new layer of indirection.

Making this design successful requires maintaining a close mapping in
the user's mind between the new representation and the original page
(cite Norman? Cognitive Dimensions?). Wildcard provides live visual cues
as the user navigates the data table, similar to the highlighting
provided by browser developer tools to indicate the mapping between HTML
and the original page.

@@

@@

Most software does not provide much user choice for interface elements,
even for common data types: e.g., the website provides a datepicker
widget, and you have no ability to provide your own replacement
datepicker. This forces users to learn many different interfaces for
similar tasks, some of them worse than others. It also prevents
combining data across multiple applications within a single user
interface.

Wildcard exposes the data underlying an application's UI, allowing the
user to use interfaces of their choice to view and modify the data. The
Expedia datepicker demo above showed one example of how this can be
useful, but we also envision creating other widgets for visualizing and
editing data. Some examples would be showing geographic data in a custom
map that includes the user's own annotations, or editing a blog post in
a rich text editor of the user's choice.

One benefit of decoupling data from interfaces is improved UI quality.
When UI widgets can compete on their merits rather than based on network
effects from the data they have access to, it creates much stronger
competition at the interface level. For example, there is competition
among email clients (which consume an open protocol), but not among
Facebook or Twitter clients. This benefit relates to the SOLID project
led by Tim Berners-Lee \autocite{berners-lee2018}, which envisions
user-controlled data as a mechanism for decoupling data from interfaces,
e.g.~giving users a choice of which client to use to consume a given
social media feed. Wildcard has overlapping goals with SOLID, but does
not require decentralized user control of data---the data can remain on
a centralized server, as long as the interface can be tweaked by end
users. \emph{Note: maybe this paragraph belongs in related work?}

There is also value in using the same consistent interface across many
applications. For example, many programmers become deeply familiar with
one text editor and use it for many different kinds of tasks. This usage
even extends beyond editing permanent files; shell programs can use the
user's preferred text editor as an interactive input mechanism (e.g.~for
editing git commit messages). This ability to generically reuse the text
editor in many contexts makes it more worthwhile to invest time
mastering the tool. Beaudouin-Lafon and Mackay refer to this type of
reuse as \emph{polymorphism} in interaction
\autocite{beaudouin-lafon2000}, noting that it is a key technique for
keeping interfaces simple while increasing their power.

\emph{Note: Maybe could relate this section to Concept Reuse?}

\emph{Note: Does this section overlap with ``expose unified
structure''?}

\hypertarget{in-place-toolchain}{%
\subsection{In-place toolchain}\label{in-place-toolchain}}

\begin{longtable}[]{@{}lll@{}}
\toprule
\endhead
& \textbf{In-place} & \textbf{Not in-place}\tabularnewline
\textbf{End user friendly} & Wildcard & IFTTT\tabularnewline
\textbf{For developers} & browser dev tools & editing OSS\tabularnewline
\bottomrule
\end{longtable}

We believe that people are more motivated to modify software when the
tools for modification are close at hand while using the original
software, referred to by Ink and Switch as an ``in-place toolchain''
\autocite{inkandswitch2019}.

Browser developer tools exemplify an in-place toolchain, since they are
built in to browsers and available in the context of viewing a page; in
contrast, creating a browser extension requires moving out of the
browser to a separate development environment.

Wildcard creates an in-place toolchain for modifying web applications.
Once installed, if an adapter is available for a site, it can be
immediately modified from within the browser. Hopefully, this can help
make switching from using to modifying software into a smooth
transition, rather than a chasm to cross.

\hypertarget{first-party-cooperation-optional}{%
\subsection{First party cooperation
optional}\label{first-party-cooperation-optional}}

The Web is an unusually extensible platform. On many other platforms
(e.g.~smartphone operating systems), software is locked down unless
first-party developers explicitly provide hooks for plugins and
interoperability, but on the Web, all client-side code is available for
browser extensions to modify. Application authors can use practices that
make it easier to modify their apps (e.g.~clean semantic markup), or
more difficult (e.g.~code obfuscation), but the default state is
openness. This gives extensions freedom to modify applications in
creative ways that the original developers did not plan for.

Wildcard takes advantage of this openness, and does not depend on
cooperation from first-party website developers. Any programmer can add
support for any website to Wildcard by building a third party adapter.
This design acknowledges the pragmatic reality that most sites today
will not have Wildcard support built in, and more broadly do not
prioritize client side extensibility.

However, we also hope that first party website developers eventually
build in Wildcard support to their applications, which would reduce the
burden of maintaining adapters and make Wildcard plugins more stable.
While this might seem optimistic, implementing the Wildcard adapter API
could be straightforward in a typical client-side application that
already has access to a structured version of the data in the page.
There is also precedent for first parties implementing an official
client extension API in response to user demand: for several years,
Google exposed an extension API in Gmail for Greasemonkey scripts to
use. (Since then, third parties have created well-maintained
alternatives \autocite{streak,talwar2019}, illustrating the value of
sharing even third party adapters.)

@@

\emph{Note: a lot of this was already covered above; how to deal with
that?}

\begin{itemize}
\tightlist
\item
  Instrumental interaction
\item
  Web automation
\item
  Wrapper induction
\item
  Personal data ownsership / SOLID
\item
  Extension helper libraries, e.g.~Gmail.js.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  still in early development; note the beta release plan (tentative:
  target public beta availability at the workshop in March?)
\item
  What are the limits of functionality? What can't it do?

  \begin{itemize}
  \tightlist
  \item
    Should we extend further? Automation workflows? Triggers?
  \end{itemize}
\item
  Could explore automated wrapper induction, building on prior work
\item
  Usability studies
\item
  Link to sign up for future updates
\end{itemize}
